### 2024-06-19 basic-statistics.qmd
20240618 - NHST

### Null Hypothesis ($H_0$)

@@

### Alternative Hypothesis ($H_1$)

@@

### Statistical Significance {#sec-statisticalSignificance}

@@

A two-by-two confusion matrix for null-hypothesis significance testing is in @fig-nhstConfusionMatrix.

@@

>>>>

### Null Hypothesis Significance Testing {#sec-nhst}

#### Null Hypothesis ($H_0$) {#sec-nullHypothesis}

@@

#### Alternative Hypothesis ($H_1$) {#sec-alternativeHypothesis}

@@

#### Statistical Significance {#sec-statisticalSignificance}

@@

A two-by-two confusion matrix for null-hypothesis significance testing is in @fig-nhstConfusionMatrix.
There are four general possibilities:

1. We can (correctly) reject the null hypothesis when it is in fact false ($1 - \beta$).
This is a true positive.
For instance, we may correctly determine that Quarterbacks have longer careers than Running Backs.
1. We can (correctly) fail to reject the null hypothesis when it is in fact true ($1 - \alpha$).
This is a true negative.
For instance, we may correctly determine that Quarterbacks do not have longer careers than Running Backs.
1. We can (incorrectly) reject the null hypothesis when it is in fact true ($\alpha$).
This is a false positive.
When performing a null hypothesis testing, a false positive is known as a Type I error.
For instance, we may incorrectly determine that Quarterbacks have longer careers than Running Backs when, in fact, Quarterbacks and Running Backs do not differ in their career length.
1. We can (incorrectly) fail to reject the null hypothesis when it is in fact false ($\beta$).
This is a false positive.
When performing a null hypothesis testing, a false negative is known as a Type II error.
For instance, we may incorrectly determine that Quarterbacks and Running Backs do not differ in their career length when, in fact, Quarterbacks have longer careers than Running Backs.

@@

A plot of statistical power is in @fig-nhst.
In statistics, *power* is the probability of detecting an effect, if, in fact, the effect exists.
Otherwise said, power is the probability of rejecting the null hypothesis, if, in fact, the null hypothesis is false.
Power is influenced by several variables:

- the sample size (*N*): the larger the *N*, the greater the power
- the [effect size](sec-practicalSignificance): the larger the effect, the greater the power
    - for group comparisons:
        - larger between-group variance, and
        - smaller within-group variance (i.e., strong measurement precision)
- the alpha level (typically .05): the higher the alpha level, the greater the power; however, the higher we set the alpha level, the higher the likelihood of Type I errors (false positives)
- one- versus two-tailed tests: one-tailed tests have higher power than two-tailed tests
- within-subject versus between-subject comparisons: within-subject designs tend to have greater power than between-subject designs

### 2024-06-19 research-methods.qmd
20240616 - placeholders for code
20240616 - Kelvin temp
20240618 - NHST

@@

- height
- weight
- age
- distance
- speed
- volume
- time elapsed
- income
- years of formal education
- points in football

>>>>

Within-subject designs tend to have greater statistical power than between-subject designs.
However, within-subject designs often have *carryover effects*.
For instance, consider the study in which we assign players to drink only water in the first and third quarters and to drink sports drink in the second and fourth quarters (an A-B-A-B design).
Drinking sports drink in the second quarter could increase how much hydration a player has throughout the rest of the game, which could lead to altered performance in the third and fourth quarters that is not due to what they drink in third and fourth quarters.

@@

- temperature on the Kelvin scale
- height
- weight
- age
- distance
- speed
- volume
- time elapsed
- income
- years of formal education
- points in football
